Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2022-07-15 sports_analytics.Rmd
tweaks to EPV section

In basketball, @cervone2014pointwise and @cervone2016multiresolution introduce expected possession value (EPV) as means toward an assessment of a player's on-court performance. 
This metric is a continuous-time estimate of the expected number of points for the offensive team on a given possession using player and ball locations. 
The EPV takes into account all possible outcomes (a shot attempt, a pass, etc.) for a given player with the ball, with different weights being assigned to each decision. 
The computation of the EPV statistic is done using a Markov model.
Consequently, the authors derive a metric called EPV-Added (EPVA), measuring a player's EPV contribution in a given situation relative to a league-average player.

A demonstration of the EPV model presented in @cervone2016multiresolution's paper is available on the first author's GitHub at <https://github.com/dcervone/EPVDemo>. 
As an illustration, using the provided tracking data, Figure \ref{fig:epv-demo} displays a snapshot of a possession during the 2013-14 NBA regular season matchup between the Miami Heat and the Brooklyn Nets on November 1, 2013. Miami is the team on offense in this play, and the possession outcome is a 26-foot three-point miss by Mario Chalmers of the Heat. 
The plot consists of two elements: 
1) (bottom) the player locations on the court at a particular moment in this possession--when the ball just left Chalmers' hands, 
and 2) (top) a line graph showing how the EPV is continuously changing throughout the play until the three-point attempt. For this entire possession, the estimated EPV for the Miami Heat reaches its peak at $1.276$ at this specific time point.

<!-- (include code demo? https://github.com/dcervone/EPVDemo) -->
<!-- > It might be cool to include one of their graphics here, if can. Or regenerate one.  -->

@@

>>>>

In basketball, @cervone2014pointwise and @cervone2016multiresolution introduce expected possession value (EPV) as a means toward an assessment of a player's on-court performance. 
This metric is a continuous-time estimate of the expected number of points for the offensive team on a given possession using player and ball locations. 
The EPV takes into account all possible outcomes (a shot attempt, a pass, etc.) for a given player with the ball, with different weights being assigned to each decision. 
The computation of the EPV statistic is done using a Markov model.
Consequently, the authors derive a metric called EPV-Added (EPVA), measuring a player's EPV contribution in a given situation relative to a league-average player.

A demonstration of the EPV model presented in @cervone2016multiresolution is available at <https://github.com/dcervone/EPVDemo>. 
Figure \ref{fig:epv-demo} illustrates how the provided tracking data informs the evolution of EPV throughout the play. 
It displays a snapshot of a possession during the NBA regular season matchup between the Miami Heat and the Brooklyn Nets on November 1, 2013. 
Miami is the team on offense in this possession, whose outcome is a 26-foot three-point miss by Mario Chalmers. 
The plot consists of two elements: 
1) (bottom) the player locations on the court at a particular moment in this possession: when the ball just left Chalmers' hands, 
and 2) (top) a line graph showing how the EPV changes continuously throughout the play until the three-point attempt. 
For this possession, the estimated EPV for the Miami Heat reaches its peak at $1.276$ points at the moment the shot is taken.

@@

Note that Miami starts the play with an EPV of approximately 1.0 points.
This indicates their implied average points per possession. 
Chalmers's shot is worth 3 points, so the EPV of 1.276 points implies that the model estimate of the probability of Chalmers making this shot is 42.5\%. 
A breakthrough in this work is that this estimate is conditional on the location of the other 9 players on the court!
